Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-28 paper/main.tex
rewrite initial applications text

\paragraph{Direct Comparison}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results.
Each cell contains a collection of vertices.
Given $M$ cells with $N$ vertices each and dimension $d$, the total information can be stored in an
array $\nu_{i,j,k}$ of shape $(M,N,2)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays within a suitable cost function~\cite{Wang2020}.
In our approach we utilize a least squares approach given by

@@

\paragraph{Testing the Comparison Algorithm with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}
\textit{Bis hierhin bin ich mit meiner Überarbeitung gekommen.}

%###################################################################################################
\section{Application 1: Elastic deformations of \textit{E.Coli}}
\label{section:application-1}

We will use our previously developed computational model and apply it directly to the data provided
by the experiments of Amir et. al~\cite{Amir2014,Amir2014_2}.
\textbf{[Es wäre gut zu sagen, was für ein Ziel du verfolgst. Warum machst du das?]}
They studied how a single \textit{E.Coli}, which is fixed at the bottom, reacts to a horizontal
fluid flow that exerts a force onto the bacteria.
This setup is displayed in figure~\ref{fig:amir-bending-simulation} (A).
To model this configuration, the lowermost vertex of the rod is fixed at the base of the simulation
domain.

Furthermore, we introduce a force $\vec{F}_i = C |\vec{x}_i - \vec{x}_{i+1}|\sin(\alpha_i)\vec{e}_x$
that acts similarly to Stoke's Law \textbf{TODO Citation} onto the normal component of each
rod-segment where $\alpha_i$ is the angle of the rod edge between $\vec{x}_i$ and $\vec{x}_{i+1}$
and the x-axis.
We extract the position of the rod from the provided image and use it as initial value.
Our simulation is split into two phases, one with an external force that initially bends the rod,
and a second one during which the force is disabled and the cell returns to its original shape.
We compare the simulated positions with the extracted ones and calculate the euclidean distance
between them, which we use as the cost function to estimate the parameters of this model.

Since this model is void of any interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$.
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the parameters for the dragging force, growth rate and damping can be identified, the same
can not be said for the rod rigidity and spring tension, which both are not identifiable.
However this behaviour is expected since, the shape of the rod is almost exclusively straight, thus
adding no further information except that a lower threshold of rigidity and spring tension should be
present.
\textbf{[Ist mir nicht ganz klar]}

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{0.38\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.62\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        Subfigure (A) shows a snapshot of the experiments performed by \cite{Amir2014_2}.
        Division of the bacteria is inhibited and they are trapped within a mother machine such that
        a controlled flow can be exerted on them.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        Subfigures (C-G) show likelihood profiles for the individual parameters.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

%###################################################################################################
\section{Application 2: \textit{E.Coli.} growing on Agar}
\label{section:application-2}
To show the usefulness of our previously developed methods
(see section~\ref{section:parameter-estimation}), we will use them to estimate the parameters
individual growing \textit{E.Coli} using a sequence of microscopic images.
The data we use~\cite{https://doi.org/10.3203/iwf/k-129} (see
section~\ref{section:data-code-availability}) has been automatically segmented with the
omnipose~\cite{Cutler2022} algorithm by mean of which the necessary cell masks are obtained.

@@

We have already suspected (see table~\ref{table:simulation-aspects}) that in order to properly
estimate parameters of our system, we need to treat every cell individually, thus assigning
parameters on a single-cell level.
This means that we increase the total number of parameters that need to be sampled within an
optimization routine.
Given $N_c$ cells and $N_p$ parameters per cell and $N_u$ individual parameters (where $N_u\leq N_p$), the total amount of parameters to be estimated is:
\begin{equation}
    N_{p,\text{total}} = N_u + N_c (N_p - N_u) = N_C N_p - (N_c - 1) N_u.
\end{equation}
It is clear, that we should avoid to needlessly assign every parameter individually since the volume
of the parameter space that needs to be sampled, scales exponentially with the dimension ie. the
number of parameters.
\textbf{[Das ist leider komplett unverständlich. Ich glaube, ich weiss schon, was du sagen möchtest, aber das muss umformuliert werden. Stichworte sind hier: Cell Variability, Heterogeneity, latent variables. Parameter per cell and individual parameters versteht auch niemand den Unterschied.]}

@@

where $\lambda$ is the growth rate and $t_0$ the delay.
Fitting this function to the individual growth curves of the bacteria and their mean growth curve,
both results in good agreements (Figure~\ref{fig:estimated-growth-rates} A,B).
However, the latter exhibits large amounts of uncertainty within the data which is also illustrated
in subfigure (C).
There, we can see much more clearly the difference between considering individually assigned
parameters and global ones.
It is clear from this observation that in order for us to accurately estimate the parameters of this
system, we need to treat the bacteria individually.
\textbf{[Man ahnt was du meinst, doch leider ahnt man es nur. Vermeide Aussagen wie "it is clear".]}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
In section we will focus on a time-interval which does not contain any division events and at which
the initial lag-phase of our bacteria has already passed.
\textbf{[Sprachlich wieder fragwürdig. "A time interval at which the initial lag-phase has passed" ist kein irgendwie korrekter englischer Satz.]}

By doing so, we can utilize the simpler and more performant direct comparison scheme for our cost function.
\textbf{[Hier weiss man nicht worüber du sprichst.]}

Furthermore, we need not to estimate the lag-time parameters which improves performance of our
optimization algorithm.
\textbf{[Verweise auf die entsprechende Gleichung.]}
\textbf{[Ich werde nun nicht mehr das Sprachliche kommentieren oder verbessern. Zuerst muss die gesamte Konzeption verbessert werden.]}

@@

With this work we have layed out how to estimate the parameters of \acp{abm} using
microscopic image time series.
\textbf{
    Es wird leider überhaupt nicht deutlich, was die Blaupause sein könnte.
    Wenn man das MS liesst, gewinnt man den Eindruck, du machst dies und das, aber ein einheitliches Konzept fehlt.
    Wir müssen zuerst eine Konzeption für das Paper entwickeln, bevor wir über weitere Details reden können.
}
\textit{Alles zu blaupause ist wegradiert.}

Our approach is modularized such that its structure can be applied to a variety of problems.
We introduced a \ac{ib} computational model for rod-shaped bacteria that incorporates bendiness of
the rod and is implemented with \texttt{cellular\_raza} allowing it to be flexible in design and
readily applicable for computational studies.
Furthermore, we devised novel methods to extract the spatial positions and other information of
cell-agents from masks which have been segmented from microscopic images.
The extracting method was validated using results from our computational model.
We further implemented a pipeline that can produce 3D visualizations of simulation snapshots and can
generate cell masks from data as produced by our computational model.
These imaging techniques were also utilized to design a cost function that can compare snapshots
of cellular states across cellular division events, allowing us to estimate parameters across a
large time-span.
Finally, the combination of these preceding advancements were applied to 2 case studies, which showed
that the profile-likelihood method can be used to effectively evaluate the ability of our system to
properly identify the involved parameters.
In the first example, we fit a simplified model of a singular bent rod to microscopic images.
These results highlighted that a reduction of the systems parameters is necessary in order to
identify them.
In our next step, the same system was fitted to a series of microscopic images in two secarios,
excluding and incorporating division events.
Both results showed the ability of our model to effectively estimate individual parameters of the
cells.
The case including division events produced parameter profiles with minima that were not fully able
to estimate the chosen parameters.
These results ca be traced back to a gap within the provided segmentation masks.

>>>>

\paragraph{Direct Comparison}
\label{paragraph:direct-comparison}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results.
Each cell contains a collection of vertices.
Given $M$ cells with $N$ vertices each and dimension $d$, the total information can be stored in an
array $\nu_{i,j,k}$ of shape $(M,N,2)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays within a suitable cost function~\cite{Wang2020}.
In our approach we utilize a least squares approach given by

@@

\paragraph{Testing the Comparison Algorithm with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}

%###################################################################################################
\section{Application 1: Elastic deformations of \textit{E.Coli}}

To assess that our model can be applied to single cells and showcase the usefulness of the
previously developed comparison techniques,
will use our previously developed computational model and apply it directly to the data provided
by the experiments of Amir et. al~\cite{Amir2014,Amir2014_2}.
They studied how a single \textit{E.Coli}, which is partially confined within a narrow tube, reacts
to a horizontal fluid flow that exerts a force onto the bacterium.
The experiment is particularly suited as a first attempt since it excludes cell-cell interactions
and focusses on the behaviour of a single cell.
It therefore reduces the total amount of parameters and allows for fast iteration times when
performing parameter estimations.
The experimental setup is displayed in figure~\ref{fig:amir-bending-simulation} (A).
To model this configuration, the lowermost vertex of the rod is fixed at the base of the simulation
domain and vertices which are inside the tubulus are restricted to move only in the y-direction.

\textbf{@Christian is this more clear now?}

We model the force exerted by the fluid flow onto the individual vertices by

\begin{equation}
    \vec{F}_i = C |\vec{x}_i - \vec{x}_{i+1}|\sin(\alpha_i)\vec{e}_x
\end{equation}

which acts similarly to Stoke's Law \textbf{TODO Citation} onto the normal component of each
rod-segment where $\alpha_i$ is the angle of the rod edge between $\vec{x}_i$ and $\vec{x}_{i+1}$
and the x-axis.
We extract the position of the rod from the provided image and use it as initial value.
Our simulation is split into two phases, one with an external force that initially bends the rod,
and a second one during which the force is disabled and the cell returns to its original shape.
We compare the simulated positions with the extracted ones and calculate the euclidean distance
between them, which we use as the cost function to estimate the parameters of this model.

Since this model is void of any cell-cell interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$ (see table~\ref{table:list-of-parameters}).
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the parameters for the dragging force, growth rate and damping can be identified, the same
can not be said for the rod rigidity and spring tension, which both are not identifiable.
However this behaviour is expected since, the shape of the rod is almost exclusively straight, thus
adding no further information except that a lower threshold of rigidity and spring tension should be
present.
\textbf{[Ist mir nicht ganz klar]}

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{0.38\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.62\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        \textbf{TODO remove pixels in units}
        \textbf{@Christian is it okay to change the units here to seconds from minutes? (compare
        table above)}
        Subfigure (A) shows a snapshot of the experiments performed by \cite{Amir2014_2}.
        Division of the bacteria is inhibited and they are partially trapped within a mother machine
        such that a controlled flow can be exerted on them.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        Subfigures (C-G) show likelihood profiles for the individual parameters.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

%###################################################################################################
\section{Application 2: \textit{E.Coli.} growing on Agar}
\label{section:application-2}
To show the usefulness of our previously developed methods
(see section~\ref{section:parameter-estimation}), we will use them to estimate the parameters of
individually growing \textit{E.Coli} using a sequence of microscopic images.
The data we use~\cite{https://doi.org/10.3203/iwf/k-129} (see
section~\ref{section:data-code-availability}) has been automatically segmented with the
omnipose~\cite{Cutler2022} algorithm to obtain the required cell masks.

@@

We have already suspected (see Table~\ref{table:simulation-aspects}) that in order to properly
estimate parameters of our system, we need to account for natural cell-variability, thus assigning
parameters on a per-cell basis.
We have previously presented a list of $N_p$ parameters (see Table~\ref{table:list-of-parameters})
which will either be assigned individually or globally.
We filter out a subset $N_g\leq N_p$ of global parameters, treating the remaining ones individually.
Given $N_c$ cells, the total amount of parameters to be estimated is:
\begin{equation}
    N_{p,\text{tot}} = N_g + N_c (N_p - N_g).
\end{equation}
It is clear, that we should avoid, to needlessly treat parameters non-globally since this increases
the size of the parameter space considerably.
The volume that needs to be sampled to optimize within a given parameter space scales exponentially
with the dimension ie. the total number of parameters $N_{p,\text{tot}}$.
\textbf{[Das ist leider komplett unverständlich. Ich glaube, ich weiss schon, was du sagen möchtest, aber das muss umformuliert werden. Stichworte sind hier: Cell Variability, Heterogeneity, latent variables. Parameter per cell and individual parameters versteht auch niemand den Unterschied.]}
\textit{Ich habe versucht, es klarer zu machen. Bitte dazu Feedback!}

@@

where $\lambda$ is the growth rate and $t_0$ the delay.
Fitting this function to the individual growth curves of the bacteria and their mean growth curve,
both results in good agreements (Figure~\ref{fig:estimated-growth-rates} A,B).
However, the latter exhibits large amounts of uncertainty within the data which is also illustrated
in subfigure (C).
There, we can see much more clearly the difference between considering assigning parameters per cell
and global ones.
These results show that in order to accurately estimate the parameters of this system, we need to
account for this inherent heterogeneity and hence treat the growth parameters of the bacteria
individually.
\textbf{[Man ahnt was du meinst, doch leider ahnt man es nur. Vermeide Aussagen wie "it is clear".]}
\textit{Jetzt klarer?}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
In this section we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant direct comparison scheme (see
\ref{paragraph:direct-comparison}) for our cost function.
We have chosen said interval such that effects related to the lag phase can be neglected which
allows us to sample a lower amount of paremeters thus improving runtime performance of our
optimization scheme.

@@

With this work we have layed out how to estimate the parameters of \acp{abm} using
microscopic image time series.
Our approach is modularized such that its structure can be applied to a variety of problems.
We introduced a \ac{ib} computational model for rod-shaped bacteria that incorporates bendiness of
the rod and is implemented with \texttt{cellular\_raza} allowing it to be flexible in design and
readily applicable for computational studies.
Furthermore, we devised novel methods to extract the spatial positions and other information of
cell-agents from masks which have been segmented from microscopic images.
The extracting method was validated using results from our computational model.
We further implemented a pipeline that can produce 3D visualizations of simulation snapshots and can
generate cell masks from data as produced by our computational model.
These imaging techniques were also utilized to design a cost function that can compare snapshots
of cellular states across cellular division events, allowing us to estimate parameters across a
large time-span.
Finally, the combination of these preceding advancements were applied to 2 case studies, which showed
that the profile-likelihood method can be used to effectively evaluate the ability of our system to
properly identify the involved parameters.
In the first example, we fit a simplified model of a singular bent rod to microscopic images.
These results highlighted that a reduction of the systems parameters is necessary in order to
identify them.
In our next step, the same system was fitted to a series of microscopic images in two secarios,
excluding and incorporating division events.
Both results showed the ability of our model to effectively estimate individual parameters of the
cells.
The case including division events produced parameter profiles with minima that were not fully able
to estimate the chosen parameters.
These results ca be traced back to a gap within the provided segmentation masks.
